---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# claimsdb

<!-- badges: start -->

[![CRAN
status](https://www.r-pkg.org/badges/version/claimsdb)](https://CRAN.R-project.org/package=claimsdb)

<!-- badges: end -->

**claimsdb** provides easy access to a sample of health insurance
enrollment and claims data from the [CMS Data Entrepreneurs' Synthetic
Public Use File
(DE-SynPUF)](https://www.cms.gov/Research-Statistics-Data-and-Systems/Downloadable-Public-Use-Files/SynPUFs/DE_Syn_PUF),
as a set of relational tables or as an in-memory database using
[DuckDB](https://duckdb.org). This package is inspired by and based on
the [starwarsdb](https://github.com/gadenbuie/starwarsdb) package.

The data are structured as actual Medicare claims data but are fully
"synthetic," after a process of alterations meant to reduce the risk of
re-identification of real Medicare beneficiaries. The synthetic process
that CMS adopted changes the co-variation across variables, so analysts
should be cautious about drawing inferences about the actual Medicare
population.

The data included in **claimsdb** comes from 500 randomly selected 2008
Medicare beneficiaries from Sample 2 of the DE-SynPUF, and it includes
all the associated claims for these members for 2008-2009. CMS provides
[resources](https://www.cms.gov/Research-Statistics-Data-and-Systems/Downloadable-Public-Use-Files/SynPUFs/DESample02),
including a codebook, FAQs, and other documents with more information
about this data.

[![Source: CMS Linkable 2008–2010 Medicare
DE-SynPUF](man/figures/README-diagram.PNG "Source: CMS Linkable 2008–2010 Medicare DE-SynPUF")](https://www.cms.gov/Research-Statistics-Data-and-Systems/Downloadable-Public-Use-Files/SynPUFs/Downloads/SynPUF_DUG.pdf)

**Formats:** [Metadata](#health-insurance-claims-data), [Local
Tables](#local-tables), [Remote Database
Tables](#remote-database-tables)

## Installation

You can install the development version of **claimsdb** from
[GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("jfangmeier/claimsdb")
```

## Health Insurance Claims Data

```{r echo=FALSE, warning=FALSE}
library(tibble)
```

All of the tables are available when you load **claimsdb**

```{r message=FALSE, warning=FALSE}
library(dplyr)
library(claimsdb)
```

The tables are also available by loading just the data from **claimsdb**

```{r eval=FALSE}
data(package = "starwarsdb")
```

This includes a `schema` table that describes each of the tables and the
included variables from the [CMS
DE-SynPUF](https://www.cms.gov/Research-Statistics-Data-and-Systems/Downloadable-Public-Use-Files/SynPUFs/DE_Syn_PUF).

```{r}
schema
```

```{r}
schema %>% 
  filter(TABLE == "inpatient") %>% 
  pull(PROPERTIES)
```

## Local Tables

Using the sample data in the tables, you can ask questions such as,
<<<<<<< HEAD
*what were the average and median prescription drug costs for men and
women in 2008?*
=======
*what were average and median prescription drug costs for men and women
in 2008?*
>>>>>>> b34b544ee71481767eec248b24c25db0e5745fa5

```{r}
rx_costs <- pde %>% 
  mutate(BENE_YEAR = lubridate::year(SRVC_DT)) %>%  
  group_by(BENE_YEAR, DESYNPUF_ID) %>% 
  summarize(TOTAL_RX_COST = sum(TOT_RX_CST_AMT, na.rm = T), .groups = "drop")

bene %>% 
  transmute(
    BENE_YEAR,
    DESYNPUF_ID,
    BENE_SEX_IDENT = case_when(
      BENE_SEX_IDENT_CD == "1" ~ "Male",
      TRUE ~ "Female")
  ) %>% 
  left_join(
    rx_costs, by = c("BENE_YEAR", "DESYNPUF_ID")
  ) %>% 
  mutate(TOTAL_RX_COST = ifelse(is.na(TOTAL_RX_COST), 0, TOTAL_RX_COST)) %>% 
  group_by(BENE_SEX_IDENT) %>% 
  summarize(
    MEAN_RX_COST = mean(TOTAL_RX_COST, na.rm = T),
    MEDIAN_RX_COST = median(TOTAL_RX_COST, na.rm = T))
```

## Remote Database Tables

Many organizations store claims data in a remote database, so
**claimsdb** also includes all of the tables as an in-memory DuckDB
database. This can be a great way to practice working with this type of
data, including building queries with dplyr code using
[dbplyr](https://dbplyr.tidyverse.org/).

```{r}
con <- claims_connect()
bene_rmt <- tbl(con, "bene")
pde_rmt <- tbl(con, "pde")
pde_rmt

rx_costs_rmt <- pde_rmt %>% 
  # note below that lubridate functions do not currently work on remote databases,
  # so you need to use date/time functions appropriate for the database.
  mutate(BENE_YEAR = date_part('year', SRVC_DT)) %>%  
  group_by(BENE_YEAR, DESYNPUF_ID) %>% 
  summarize(TOTAL_RX_COST = sum(TOT_RX_CST_AMT, na.rm = T), .groups = "drop") %>% 
  ungroup()
rx_costs_rmt

bene_rmt %>% 
  transmute(
    BENE_YEAR,
    DESYNPUF_ID,
    BENE_SEX_IDENT = case_when(
      BENE_SEX_IDENT_CD == "1" ~ "Male",
      TRUE ~ "Female")
  ) %>% 
  left_join(
    rx_costs_rmt, by = c("BENE_YEAR", "DESYNPUF_ID")
  ) %>% 
  mutate(TOTAL_RX_COST = ifelse(is.na(TOTAL_RX_COST), 0, TOTAL_RX_COST)) %>% 
  group_by(BENE_SEX_IDENT) %>% 
  summarize(
    MEAN_RX_COST = mean(TOTAL_RX_COST, na.rm = T),
    MEDIAN_RX_COST = median(TOTAL_RX_COST, na.rm = T))
```
